---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This report is written as part of the John Hopkins course on Reproducable Research
presented on Coursera. The aim of the assignment is to demonstrate how a large 
data set can be utilized in reproducible research by using literate programming. 
The report is produced using r markdown to present both the r code and the 
results of that code together with the written text into one file that is 
knitted into an html file.

A dataset has been provided as part of the assignment. The data was collected from
a personal activity monitoring device. During the period of October and November
in 2012 the steps taken by an anonymous individual were recorded in 5 minute 
intervals throughout each day.


## Loading and preprocessing the data

The data needed for the assignment is stored in the [github][1] repository
as a zip file. 


**Obtaining the data from github**

The first step is to download the zip file into a temporary folder. 
From the temporary folder the .csv file is extracted using unz() to unzip the file.


```{r loadingGit}
    
    dataTemp <- tempfile()

    url1 <- "https://github.com/sandjinn/RepData_PeerAssessment1/blob/master/activity.zip?raw=TRUE"
    
    download.file(url1, dataTemp, mode = "wb")
    
    dataCSV <- read.csv(unz(dataTemp, "activity.csv"))
    
    unlink(dataTemp)
```


## What is mean total number of steps taken per day?

The steps for each day are recorded in 5 minute intervals. In order to calculate
the average (in this case the mean) number of steps per day the total number of steps
taken each day have to be calculated first. 


**Calculate the total number of steps taken per day**

By using the tapply() function the total number of steps per day can be 
calculated using sum.

```{r stepsDay}

    stepsSum <- tapply(dataCSV$steps, dataCSV$date, sum)
```


**Make a histogram of the total number of steps taken each day**

A histogram of the total number of steps taken each day (calculated above) shows 
a fairly normal distribution.

```{r stepsHist}

    hist(stepsSum, main="Histogram of daily step count", 
         xlab="Total number of steps taken per day")

```


**Calculate and report the mean and median of the total number of steps/day**

```{r stepMean}

    smean <- mean(stepsSum, na.rm = TRUE)

```

```{r stepMedian}

    smedian <- median(stepsSum, na.rm = TRUE)

```

The mean value of the total number of steps taken per day is `r smean`.

The median of the total number of steps taken per day is `r smedian`.



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

[1]: https://github.com/rdpeng/RepData_PeerAssessment1
